---
title: Keep or Drop Selected Variables
author: "Shu Fai Cheung & Mark Hok Chio Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Keep or Drop Selected Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  =  6,
  fig.height =  6,
  fig.align = "center"
)
```

```{r setup, echo = FALSE}
library(semptools)
```

# Introduction

In psychological research, it is common for a path analytic model to have
several control variables. They need to be included in the analysis. However,
for generating the plot, it is acceptable to omit them as long as results
regarding these control variables are reported in other forms (e.g., a table
of all parameter estimates), and keep only the main variables in the plot.

Two helper functions, `drop_nodes` and `keep_nodes`, were developed for
this purpose. When an SEM output object, such as the output from `lavaan`,
is passed to `semPlot::semPaths`, it is first processed by
`semPlot::semPlotModel`. In other words, `semPlot::semPaths` itself
does not interpret the SEM output. It actually use the output of
`semPlot::semPlotModel` to generate the plot.
These helper functions modify the output
of `semPlot::semPlotModel` to be used by `semPlot::semPaths`.

# Example

Suppose we have a model with four main variables, `x1`, `x2`, `x3`, and
`x4`, and three control variables, `cov1`, `cov2`, and `cov3`. The
sample data, `pa_example_3covs`, is in the package `semptools`.

This is the analysis in `lavaan`:

```{r}
library(lavaan)
mod <-
  'x3 ~  x1 + x2 + cov1 + cov2 + cov3
   x4 ~  x1 + x3 + cov1 + cov2 + cov3
  '
fit <- lavaan::sem(mod, pa_example_3covs)
```

We can plot the full model using `semPlot::semPaths` and
use `layout` and `layout_matrix` to control the positions
of all seven variables (please refer to the vignette on
`layou_matrix` on how to do this):

```{r}
library(semPlot)
library(semptools)
m <- layout_matrix(x1 = c(1, 1),
                   x2 = c(3, 1),
                   x3 = c(2, 2),
                   x4 = c(2, 3),
                   cov1 = c(4, 1),
                   cov2 = c(5, 1),
                   cov3 = c(6, 1))
p_pa <- semPaths(fit, whatLabels = "est",
           sizeMan = 10,
           edge.label.cex = .5,
           style = "ram",
           nCharNodes = 0, nCharEdges = 0,
           layout = m)
```

Suppose we want to remove `cov1`, `cov2`, and
`cov3`. We do this by calling `semPlot::semPlotModel`
directly, and modify it by `keep_nodes` or `drop_nodes`.

We can *drop* `cov1`, `cov2`, and `cov3` by `drop_nodes`:

```{r}
pm_no_covs <- semptools::drop_nodes(
                object = semPlotModel(fit),
                nodes = c("cov1", "cov2", "cov3"))
```

The first argument, `object`, should be the output of
`semPlot::semPlotModel`. In the example, `semPlotModel(fit)`
is used to call `semPlot::semPlotModel` to process `fit`
and then pass the results immediately to `drop_nodes`.

The second argument, `nodes`, is a character vector with
the names of the variables to be dropped.

With just two arguments, the argument names can be omitted:

```{r}
pm_no_covs <- semptools::drop_nodes(
                semPlotModel(fit),
                c("cov1", "cov2", "cov3"))
```

We can then use `semPlot::semPaths` to plot this modified
model:

```{r}
m_no_covs <- layout_matrix(x1 = c(1, 1),
                           x2 = c(3, 1),
                           x3 = c(2, 2),
                           x4 = c(2, 3))
pa_no_covs <- semPaths(pm_no_covs, whatLabels = "est",
                    sizeMan = 10,
                    edge.label.cex = .5,
                    style = "ram",
                    nCharNodes = 0, nCharEdges = 0,
                    layout = m_no_covs)
```

Note that in the call to `semPlot::semPaths`, the modified
output of `semPlot::semPlotModel`, `pm_no_covs`, is used
instead of `fit`, the output of `lavaan`.

Alternatively, we can also specify the variables to *keep*
using `keep_nodes`. It is used in a similar way, except that
the variables specified in `nodes` will be kept, and all
variables *not specified* will be removed.

```{r}
pm_only_xs <- semptools::keep_nodes(
                semPlotModel(fit),
                c("x1", "x2", "x3", "x4"))
pa_only_xs <- semPaths(pm_only_xs, whatLabels = "est",
                    sizeMan = 10,
                    edge.label.cex = .5,
                    style = "ram",
                    nCharNodes = 0, nCharEdges = 0,
                    layout = m_no_covs)
```

This plot is identical to the previous plot.

The plot generated
by `semPlot::semPaths` with selected nodes dropped or
kept can then be passed to other
`semptools` functions for further processing.

# Keep Or Drop? It Depends

Which function to use depends on which one is easier
to specify. For example,
if there are a lot of contorl variables but only a few
main variables, then it is more efficient to use `keep_nodes`
and specify the few main variables. If the number of
control variables is substantially less than the number of
main variables, then it is more efficient to use `drop_nodes`
and specify only the few control variables to drop.