---
title: Layout Matrices  
author: "Shu Fai Cheung & Mark Hok Chio Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{layout_matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  =  6,
  fig.height =  6,
  fig.align = "center"
)
```

```{r setup, echo = FALSE}
library(semptools)
```

# Introduction

One useful feature of `semPlot::semPaths` is using setting the argument `layout`
to a layout matrix to specify the location of each node in a structural
equation model. This guide briefly explains how to use `layout`, and then introduces
the helper function `layout_matrix` of `semptools` that can be used to construct
a layout matrix. Last, it describes how `factor_layout` is used in
`set_sem_layout`
to specify the layout of the latent factors only, and let `set_sem_layout`  
decide the positions of their indicators.

# What `layout` does in `semPlot::semPaths`.

Suppose we have a path model with four variables, 
`x1`, `x2`, `x3`, and `x4`. `x1` and `x2` affects `x3`,
and `x1` and `x3` affects `x4`. In psychology, this
model is usually presented this way:

```{r conceptual_diagram, echo = FALSE}
suppressMessages(library(lavaan))
suppressMessages(library(semPlot))
mod_pa <- 
 'x1 ~~ x2
  x3 ~  x1 + x2
  x4 ~  x1 + x3
 '
fit_pa <- lavaan::sem(mod_pa, pa_example)
m <- matrix(c("x1",   NA,  NA,   NA,
                NA, "x3",  NA, "x4",
              "x2",   NA,  NA,   NA), byrow = TRUE, 3, 4)
p_pa <- semPaths(fit_pa, whatLabels = "path",
           sizeMan = 10,
           edge.label.cex = 1.15,
           style = "ram",
           nCharNodes = 0, nCharEdges = 0,
           layout = m,
           residuals = FALSE,
           DoNotPlot = TRUE)
p_pa2con <- set_curve(p_pa, list(list(from = "x1", to = "x2", new_curve = -2)))
plot(p_pa2con)
```

If we fit the model by `lavaan` and use `semPlot::semPaths` to generate
the plot without using `layout`, this is the resulting plot:

```{r wo_layout}
library(lavaan)
library(semPlot)
mod_pa <- 
 'x1 ~~ x2
  x3 ~  x1 + x2
  x4 ~  x1 + x3
 '
fit_pa <- lavaan::sem(mod_pa, pa_example)
p_pa <- semPaths(fit_pa, whatLabels = "est",
           sizeMan = 10,
           edge.label.cex = 1.15,
           style = "ram", 
           nCharNodes = 0, nCharEdges = 0)
```

This layout is different from the convention used in psychology.

To use `layout`, we first decide the grid to be used to 
position the variable (nodes). For example, for the
conceptual diagram, we can try a 3 by 4 grid

```{r grid, echo = FALSE}
plot(p_pa2con)
xc <- seq(-1.25, 1.25, length.out = 5)
yc <- seq(1.25, -1.25, length.out = 4)
for (i in yc) {
    segments(xc[1], i, xc[5], i, col = "red", lwd = 4)
  }
for (i in xc) {
    segments(i, yc[1], i, yc[4], col = "red", lwd = 4)
  }
x_os <- (xc[2] - xc[1]) * .5
y_os <- (yc[2] - yc[1]) * .9
for (i in seq_len(length(xc) - 1)) {
    for (j in seq_len(length(yc) - 1)) {
        text(xc[i] + x_os, yc[j] + y_os,
             paste0("(", j, ", ", i, ")"),
             cex = 1.5, col = "red")
      }
  }
```

The pair of numbers in each cell denote the location of
the cell in a 3 by 4 matrix as indexed in an R matrix.
An empty column was added between `x3` and `x4` because we
want to move `x4` further away to the right.

We then create a matrix of the same dimension as the
grid, and initialize the cells by `NA`, which denoted a cell
with *nothing*.

```{r}
m <- matrix(NA, 3, 4)
m
```

We then set the position of each variable
(node, as called internally in a plot by `semPlot::semPaths`)
by setting the corresponding cell to the name
of this variable as appeared in the `lavaan` model.

```{r}
m[1, 1] <- "x1"
m[3, 1] <- "x2"
m[2, 2] <- "x3"
m[2, 4] <- "x4"
m
```

We can then set `layout` to this matrix to tell 
`semPlot::semPaths` how to position the four variables:

```{r}
p_pa <- semPaths(fit_pa, whatLabels = "est",
           sizeMan = 10,
           edge.label.cex = 1.15,
           style = "ram", 
           nCharNodes = 0, nCharEdges = 0,
           layout = m)
```

Alternatively, we can type the matrix as it would appear
if printed, and set `byrow = TRUE`:

```{r}
m <- matrix(c("x1",   NA,  NA,   NA,
                NA, "x3",  NA, "x4",
              "x2",   NA,  NA,   NA), byrow = TRUE, 3, 4)
m
```

We need to type more because we need to include
`NA` for all the empty cells. However, 
this approach let
us see immediately how the variables will be positioned.
We just place the variables in the target cell, without
knowing the coordinates. This is a WYSIWYG
(what-you-see-is-what-you-get) approach.
This is the approach used in the Quick Start Guides.

# `layout_matrix` in `semptools`

The WYSIWYG approach in the previous
section has one drawback: It is not easy to change the
position of the variables and the dimension of the grid.
In real research, trial-and-error is usually needed to 
find a desirable layout. For example, if we want to
add a column or row,
we need ty type several `NA`s to crate it.

The `layout_matrix` function in `semptools`
is designed to generate the matrix using the coordinates
of the variables. Instead of specifying the dimension
ourselves, `layout_matrix` will try to figure out the
dimension based on the coordinates automatically.

For exmaple, to generate the same layout above, we can
do this:

```{r}
m2 <- layout_matrix(x1 = c(1, 1),
                    x2 = c(3, 1),
                    x3 = c(2, 2),
                    x4 = c(2, 4))
m2
p_pa <- semPaths(fit_pa, whatLabels = "est",
           sizeMan = 10,
           edge.label.cex = 1.15,
           style = "ram", 
           nCharNodes = 0, nCharEdges = 0,
           layout = m2)
```

Suppose we want to move `x4` closer to `x3`. Instead
of deleting the 3rd columns of `NA`, we can  
just change the coordinates of `x4` in `layout_matrix`:

```{r}
m3 <- layout_matrix(x1 = c(1, 1),
                    x2 = c(3, 1),
                    x3 = c(2, 2),
                    x4 = c(2, 3))
m3
p_pa <- semPaths(fit_pa, whatLabels = "est",
           sizeMan = 10,
           edge.label.cex = 1.15,
           style = "ram", 
           nCharNodes = 0, nCharEdges = 0,
           layout = m3)
```

The best approach to specify the layout depends on
the situation at hand. During the trial-and-error
phrase, using `layout_matrix` is good for changing
the layout. When the layout has been finalized,
for readability, typing the matrix row-by-row may
be better (although we can still use `layout_matrix`
to form the matrix and then print the matrix, as we
did above).

# `factor_layout` in `set_sem_layout`

If we
use `layout` in `semPlot::semPaths` for a structural
models with latent factors
and we want to draw both the factors and their indicators,
we need to specify the positions of *all* nodes, that is,
all the indicators and all the factors. The grid will be
very large and it is difficult to determine the positions
of the indicators.

The `factor_layout` argument in `set_sem_layout` is developed
to solve this problem. It works
like `layout` in `semPlot::semPaths`.
In `set_sem_layout`,
we only need to specify the positions of the latent
factors. As in `layout`, we create a matrix to specify
the positions of the factors. The procedure is identical
to what illustrated above for a path model, and all
the approaches presented above can be used. The only
difference is, only the names of the *latent* *factors* need
to be present in the matrix, and the size of the grid
only need to consider the factors. 

For example, suppose we have 14 indicators and four
factors, and this is the model:

```{r}
mod <- 
  'f1 =~ x01 + x02 + x03
   f2 =~ x04 + x05 + x06 + x07
   f3 =~ x08 + x09 + x10
   f4 =~ x11 + x12 + x13 + x14
   f3 ~  f1 + f2
   f4 ~  f1 + f3
  '
```

If we want to draw both the factors and the indicators,
the plot will have 18 nodes:

```{r}
fit <- lavaan::sem(mod, cfa_example)
p <- semPaths(fit, whatLabels="est",
        sizeMan = 5,
        node.width = 1,
        edge.label.cex = .75,
        style = "ram",
        mar = c(5, 5, 5, 5))
```

Suppose we want to position the factors this way:

```{r echo = FALSE}
m <- matrix(c("f1",   NA,   NA,
                NA, "f3", "f4",
              "f2",   NA,   NA), byrow = TRUE, 3, 3)
p_sem <- semPaths(fit, what = "mod",
           sizeMan = 10,
           edge.label.cex = 1.15,
           style = "ram",
           nCharNodes = 0, nCharEdges = 0,
           layout = m,
           residuals = FALSE,
           structural = TRUE,
           DoNotPlot = TRUE)
p_semcon <- set_curve(p_sem, list(list(from = "f1", to = "f2", new_curve = -2)))
plot(p_semcon)
```

Again, we decide the grid to use, which is 3 by 3 in
this example:

```{r echo = FALSE}
plot(p_semcon)
xc <- seq(-1.25, 1.25, length.out = 4)
yc <- seq(1.25, -1.25, length.out = 4)
for (i in yc) {
    segments(xc[1], i, xc[4], i, col = "red", lwd = 4)
  }
for (i in xc) {
    segments(i, yc[1], i, yc[4], col = "red", lwd = 4)
  }
x_os <- (xc[2] - xc[1]) * .5
y_os <- (yc[2] - yc[1]) * .9
for (i in seq_len(length(xc) - 1)) {
    for (j in seq_len(length(yc) - 1)) {
        text(xc[i] + x_os, yc[j] + y_os,
             paste0("(", j, ", ", i, ")"),
             cex = 1.5, col = "red")
      }
  }
```

Therefore, this is the matrix to be used:

```{r}
m_sem <- layout_matrix(f1 = c(1, 1),
                    f2 = c(3, 1),
                    f3 = c(2, 2),
                    f4 = c(2, 3))
m_sem
```

We can then use this matrix in `set_sem_layout` (please
refer to the quick start guide of `set_sem_layout` on
how to specify other arguments):

```{r}
indicator_order  <- c("x04", "x05", "x06", "x07", 
                      "x01", "x02", "x03", 
                      "x11", "x12", "x13", "x14",
                      "x08", "x09", "x10")
indicator_factor <- c( "f2",  "f2",  "f2",  "f2",  
                       "f1",  "f1",  "f1",  
                       "f4",  "f4",  "f4",  "f4",  
                       "f3",  "f3",  "f3")
factor_point_to <- matrix(c("left",     NA,      NA,
                                NA, "down",    "up",
                            "left",     NA,      NA), byrow = TRUE, 3, 3)
indicator_push <- list(list(node = "f3", push = 2.5),
                       list(node = "f4", push = 2.5),
                       list(node = "f1", push = 1.5),
                       list(node = "f2", push = 1.5))
indicator_spread <- list(list(node = "f1", spread =    2),
                         list(node = "f2", spread =    2),
                         list(node = "f4", spread =    2),
                         list(node = "f3", spread = 1.75))
loading_position <- list(list(node = "f2", position = .6),
                         list(node = "f3", position = .8),
                         list(node = "f4", position = .8))
p2 <- set_sem_layout(p, 
                     indicator_order = indicator_order,
                     indicator_factor = indicator_factor,
                     factor_layout = m_sem,
                     factor_point_to = factor_point_to,
                     indicator_push = indicator_push,
                     indicator_spread = indicator_spread,
                     loading_position = loading_position)
plot(p2)
```

This make it much easier to specify the positions of factors
in a structural equation model with latent factors.

