---
title: "A quick start guide on using set_sem_layout"
author: "Shu Fai Cheung & Mark Hok Chio Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{quick_start_sem}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  =  8,
  fig.height =  6,
  fig.align = "center"
)
```

```{r setup, echo = FALSE}
```

# Introduction

The semptools package contains functions that *post-process* an output from 
`semPlot::semPaths`, to help users to customize the appearance of the graphs 
generated by `semPlot::semPaths`. For the introduction to functions for doing 
very specific tasks, such as moving the parameter estimate of a path or rotating
the residual of a variable, please refer to [Quick Start Guide](quick_start.html).
The present guide focuses on how to use `set_sem_layout()` to configure various 
aspects of an `semPaths` graph generated for a typical structural equation model (SEM)
with latent factors. For configuring the layout of a confirmatory factor analysis (CFA)
model with no structural path between factors, please refer to the [guide for 
set_cfa_layout](quick_start_cfa.html).

# An semPaths graph of an SEM model with minimal configurations

Let us consider an SEM model. We will use `sem_example`, a sample CFA dataset 
from semptools with 14 variables for illustration.

```{r}
library(semptools)
head(sem_example)
```

This is the SEM model to be fitted:

```{r}
mod <- 
  'f1 =~ x01 + x02 + x03
   f2 =~ x04 + x05 + x06 + x07
   f3 =~ x08 + x09 + x10
   f4 =~ x11 + x12 + x13 + x14
   f3 ~  f1 + f2
   f4 ~  f1 + f3
  '
```

The lavaan results:

```{r}
library(lavaan)
fit <- lavaan::sem(mod, cfa_example)
summary(fit)
```

This is the plot from `semPaths`:

```{r}
library(semPlot)
p <- semPaths(fit, whatLabels="est",
        sizeMan = 5,
        node.width = 1,
        edge.label.cex = .75,
        style = "ram",
        mar = c(5, 5, 5, 5))
```

We will see how `set_sem_layout()` can be used to do the following tasks to *post-process*
the graph:

- Change the order of the indicators.

- Assign factors to indicators manually.

- Specify how to set the approximate positions of the factors.

- Specify how to place the indicators of a factor.

- Adjust the placement of the indicators relative to their corresponding factors.

- Move the loadings along the paths from factors to indicators.

# Specify the order of the indicators, assign them to factors, and specify the approximate positions of the factors

Suppose we want to do this:

- We would like to assign the indicators to the factors this way:

    - x04, x05, x06, and x07 for f2.
    
    - x01, x02, and x03 for f1.
    
    - x11, x12, x13, and x14 for f4.
    
    - x08, x09, and x10 for f3.
    
To do this, we create two vectors, one for the argument `indicator_order` and 
the other for the argument `indicator_factor`. 

- `indicator_order` is a string vector with length equal to the number of indicators, with the desired order if the indicators are placed *above* the corresponding factors. In this example, it will be like this:

```{r}
indicator_order  <- c("x04", "x05", "x06", "x07", 
                      "x01", "x02", "x03", 
                      "x11", "x12", "x13", "x14",
                      "x08", "x09", "x10")
```

- `indicator_factor` is a string vector with length equal to the number of indicators. The elements are the names of the latent factors, denoting which factor each indicator will be assigned to:

```{r}
indicator_factor <- c( "f2",  "f2",  "f2",  "f2",  
                       "f1",  "f1",  "f1",  
                       "f4",  "f4",  "f4",  "f4",  
                       "f3",  "f3",  "f3")
```

To specify the locations of the factors, we need two more arguments, 
`factor_layout` and `factor_point_to`. 

`factor_layout` is a matrix of arbitrary size, with either `NA` or the name of
 a factor. For example:
 
```{r}
factor_layout <- matrix(c("f1",   NA,   NA,
                           NA, "f3", "f4",
                         "f2",   NA,   NA), byrow = TRUE, 3, 3)
```

This sets up a 3-by-3 grid, with `f1` on the top left, `f2` on the bottom left,
`f3` in the center, and `f4` on the right of `f3`. Each factor must be in one
and only one cell of this matrix.

Note that a column or row can contain only `NA`, to increase the vertical or 
horizontal distance between factors.

`factor_point_to` is a matrix of the size as `factor_layout`, with either `NA` 
or one of these: "down", "left", "up", or "right", to indicate the direction
that a factor "points to" its indicator. For example:


```{r}
factor_point_to <- matrix(c("left",     NA,      NA,
                                NA, "down", "down",
                            "left",     NA,      NA), byrow = TRUE, 3, 3)
```

`f1` and `f2` will point to the left (i.e., indicators on the left), `f3` and `f4`
will point downwards.

In sum, the `set_sem_layout()` function needs at least four arguments:

- `semPaths_plot`: The `semPaths` plot.

- `indicator_order`: The vector for the order of indicators.

- `indicator_factor`: The vector for assigning indicators to latent factors.

- `factor_layout`: The position of the factors on a grid.

- `factor_point_to`: The placement of the indicators.
    
They do not have to be named if they are in this order.

We now use `set_sem_layout()` to post-process the graph:

```{r}
p2 <- set_sem_layout(p, 
                     indicator_order = indicator_order,
                     indicator_factor = indicator_factor,
                     factor_layout = factor_layout,
                     factor_point_to = factor_point_to)
plot(p2)
```

# Adjust the placement of the indicators relative to their corresponding factors.

The placement of the indicators are too close to the indicators and to neighboring indicators. We can adjust the relative position in two ways.

## "Pushing" the indicators away from their assigned factors

We can use the argument `indicator_push` to push the indicators of a factor 
away from it. The argument needs a named vector. The name is the factor
of which the indictors will be "pushed", and the value is how "hard" the
push is: the multiplier to the 
distance from the factor to the indicators. For example:

```{r}
indicator_push <- c(f3 = 2,
                    f4 = 1.5,
                    f1 = 1.5,
                    f2 = 1.5)
```

This vector will double the distance between the indicators of `f3` and their 
factors, and multiply
the distance between the indicators of `f4`, `f1`, and `f2` and their factors 
by 1.5. If `push` is less than 1, 
the indicators will be "pulled" towards their factors.

```{r}
p2 <- set_sem_layout(p, 
                     indicator_order = indicator_order,
                     indicator_factor = indicator_factor,
                     factor_layout = factor_layout,
                     factor_point_to = factor_point_to,
                     indicator_push = indicator_push)
plot(p2)
```

## "Spreading" out the indicators

We can use the argument `indicator_spread` to spread out the indicators of a factor,
increasing the distance between the indicators. The argument needs a named
vector. The name is the fator of which the indicators will be spread out.
The value is the multiplier to the distance between neighboring 
indicators. For example:

```{r}
indicator_spread <- c(f1 = 2,
                      f2 = 1.5,
                      f4 = 1.5)
```

This vector will double the distance between the indicators of `f1`, and multiply
the distance between the indicators of `f2` and `f4`. and its indicators by 
1.5. If `spread` is less than 1,  the indicators will be squeezed towards each 
others.

```{r}
p2 <- set_sem_layout(p, 
                     indicator_order = indicator_order,
                     indicator_factor = indicator_factor,
                     factor_layout = factor_layout,
                     factor_point_to = factor_point_to,
                     indicator_push = indicator_push,
                     indicator_spread = indicator_spread)
plot(p2)
```

# Move the loadings along the paths from factors to indicators.

We can move the loadings of indicators along the paths by the argument
`loading_position`. If we supply one single number, from 0 to 1, this number 
will be used for the position of all loadings. A value of .5 place the loadings
on the middle of the paths. Larger the value, closer the loadings to the indicators.
Smaller the value, closer the loadings to the factors.

We can also use a named vector to specify the positions of indicators for 
each factor.In each element, the name if the factor whose loadings
will be moved. The value is the positons of its loadings.  The default is .50. We only need to specify the positions for factors to be changed from .50
to other values. For example:

```{r}
loading_position <- c(f2 = .7,
                      f3 = .8,
                      f4 = .8)
```


```{r}
p2 <- set_sem_layout(p, 
                     indicator_order = indicator_order,
                     indicator_factor = indicator_factor,
                     factor_layout = factor_layout,
                     factor_point_to = factor_point_to,
                     indicator_push = indicator_push,
                     indicator_spread = indicator_spread,
                     loading_position = loading_position)
plot(p2)
```

# Pipe

Like other functions in `semptools`, the `set_sem_layout()` function can be chained
with other functions using the pipe operator, `%>%`, from the package `magrittr`,
or the native pipe `|>` available since R 4.1.x.
Suppose we want to mark the significant test results for the free parameters using
`mark_sig()`, and use `set_curve()` to change the curvature of `f1 ~~ f2` covariances
and `f4 ~ f1` paths (we push and spread some indicators to make room
for the asterisks, and change the orientation of `f4` to `up`):

```{r}
factor_point_to <- matrix(c("left",     NA,      NA,
                                NA, "down",    "up",
                            "left",     NA,      NA), byrow = TRUE, 3, 3)
indicator_push <- c(f3 = 2.5,
                    f4 = 2.5,
                    f1 = 1.5,
                    f2 = 1.5)
indicator_spread <- c(f1 = 2,
                      f2 = 2,
                      f4 = 2,
                      f3 = 1.75)
loading_position <- c(f2 = .6,
                      f3 = .8,
                      f4 = .8)
if ((R.version$major > 4) & (R.version$minor > 1.0)) {
  p2 <- set_sem_layout(p, 
                      indicator_order = indicator_order,
                      indicator_factor = indicator_factor,
                      factor_layout = factor_layout,
                      factor_point_to = factor_point_to,
                      indicator_push = indicator_push,
                      indicator_spread = indicator_spread,
                      loading_position = loading_position) |>
                      set_curve(c("f2 ~~ f1" = -1,
                                  "f4 ~ f1" = 1.5)) |>
                      mark_sig(fit)
  plot(p2)
}
```

# Limitations

- Currently, if a function needs the SEM output, only lavaan output is supported.