---
title: "Models with Second-Order Factors"
author: "Shu Fai Cheung & Mark Hok Chio Lai"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Models with Second-Order Factors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{magrittr}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  =  8,
  fig.height =  6,
  fig.align = "center"
)
```

# Introduction

This article illustrates how to use `set_sem_layout()`
for a model with second-order factors. Readers
are assumed to have learned how to use `set_sem_layout()`
(see `vignette("quick_start_sem")`).

# The Initial `semPaths` Plot

We will use `sem_2nd_order_example`, a sample SEM dataset
from semptools with 21 variables for illustration.

```{r}
library(semptools)
head(round(sem_2nd_order_example, 3), 3)
```

This is the SEM model to be fitted:

```{r}
mod <-
  'f1 =~ x01 + x02 + x03
   f2 =~ x04 + x05 + x06 + x07
   f3 =~ x08 + x09 + x10
   f4 =~ x11 + x12 + x13 + x14
   f5 =~ x15 + x16 + x17 + x18
   f6 =~ x19 + x20 + x21
   f21 =~ 1*f1 + f3 + f4
   f22 =~ 1*f2 + f5 + f6
   f22 ~ f21
  '
```

Fitting the model using  `lavaan::sem()`:

```{r}
library(lavaan)
fit <- lavaan::sem(mod, sem_2nd_order_example)
```

This is the plot from `semPaths`:

```{r}
library(semPlot)
p <- semPaths(fit, whatLabels = "est",
              sizeMan = 5,
              nCharNodes = 0, nCharEdges = 0,
              edge.width = 0.8, node.width = 0.7,
              edge.label.cex = 0.6,
              style = "ram",
              mar = c(5, 5, 5, 5))
```

# Modify the Plot

As described in `vignette("set_sem_layout")`, the
plot can be modified by `set_sem_layout()`, with
these special treatment:

## Special Treatment

To modify a plot with second order factors:

- The "indicators" of a second-order factor
  need to be included in the vector for
  the `indicator_order` argument.

- In the matrix for `factor_layout`, all
  factors, first or second order, should
  be included.

- Although a second-order factor has
  latent factors as their indicators,
  the direction of a second-order
  factor in the `factor_point_to`
  matrix is ignored. The position of
  the indicators of a second-order
  factor are determined by the
  `factor_layout` matrix.

In the example below, `f1` to `f6`,
though latent factors themselves,
are included in the vector for
`indicator_order`.

## Example

```{r}
indicator_order  <- c("x01", "x03", "x02",
                      "x05", "x06", "x04", "x07",
                      "x08", "x09", "x10",
                      "x12", "x11", "x13", "x14",
                      "x16", "x15", "x18", "x17",
                      "x20", "x19", "x21",
                      "f1", "f3", "f4",
                      "f5", "f6", "f2")
indicator_factor <- c("f1", "f1", "f1",
                      "f2", "f2", "f2", "f2",
                      "f3", "f3", "f3",
                      "f4", "f4", "f4", "f4",
                      "f5", "f5", "f5", "f5",
                      "f6", "f6", "f6",
                      "f21", "f21", "f21",
                      "f22", "f22", "f22")
factor_layout <- matrix(c(  NA, "f21",   NA, NA,  NA, "f22",   NA,
                          "f1",  "f4", "f3", NA, "f2",  "f6", "f5"),
                          byrow = TRUE, 2, 7)
factor_layout <- matrix(c("f1",    NA,    NA, "f2",
                          "f4", "f21", "f22", "f6",
                          "f3",    NA,    NA, "f5"),
                          byrow = TRUE, 3, 4)
factor_point_to <- matrix(c("left",     NA,      NA, "right",
                            "left", "left", "right", "right",
                            "left",     NA,      NA, "right"),
                            byrow = TRUE, 3, 4)
indicator_spread <- c(f4 = 1.25,
                      f2 = 1.25,
                      f5 = 1.25)
p2 <- set_sem_layout(p,
                     indicator_order = indicator_order,
                     indicator_factor = indicator_factor,
                     factor_layout = factor_layout,
                     factor_point_to = factor_point_to,
                     indicator_spread = indicator_spread)
```

This is the result:

```{r}
plot(p2)
```

The layout is acceptable. It can be further modified
by other functions described in `vignette("semptools")`.
For example, the residuals can be rotated:

```{r}
my_rotate_resid_list <- c(f4  =  45,
                          f21 =   0,
                          f22 =   0,
                          f6  = -45)
p3 <- rotate_resid(p2, my_rotate_resid_list)
plot(p3)
```

# Further information

Please refer to `vignette("quick_start_sem")`
on other options of `set_sem_layout()` and
how to use it with other functions in the package.