---
title: "Using set_sem_layout() for a Model With Observed Exogenous Variables"
author: "Shu Fai Cheung & Mark Hok Chio Lai"
output:
  html_document:
    fig.align: "center"
    toc: true
    number_sections: false
bibliography: references.bib
csl: apa.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width  =  8,
  fig.height =  6,
  fig.align = "center"
)
```

# Goal

This article illustrates how to use
`set_sem_layout()` from the package
[semptools](https://sfcheung.github.io/semptools/)
([CRAN page](https://cran.r-project.org/package=semptools))
for a model with
latent variables and one or more
observed exogenous variables (a.k.a.,
predictors or independent variables
in some disciplines), and
the observed exogenous variables to be
placed like other latent variables
in a model diagram.

This article assumes basic knowledge
in using `semptools` and `set_sem_layout()`
(see `vignette("semptools")`
and `vignette("quick_start_sem")` for
illustrations).

# Model and Data

The following model is to be fitted
to the dataset `sem_example`, provided
in the package `semptools`. It is
fitted by `lavaan::sem()`:

```{r}
library(semptools)
library(lavaan)
mod <-
  'f1 =~ x01 + x02 + x03
   f3 =~ x08 + x09 + x10
   f4 =~ x11 + x12 + x13 + x14
   f3 ~ f1 + x04
   f4 ~ f3 + x05'
fit_sem <- lavaan::sem(mod, sem_example)
```

In this model, `x04` and `x05` are
exogenous observed variables. They
affect `f3` and `f4` but they themselves
are not indicators of latent factors.

# Setting `indicator_order` and `indicator_factor`

For a model in which all observed variables
are indicators of latent factors, they
are placed in the vector for `indicator_order`.

If an observed variable is not an indicator,
then this is how to include it in
the call to `set_sem_layout()`:

- It should be placed in `indicator_order`
  and `indicator_factor` as if it were
  a latent factor and it were its own
  indicator.

- In the matrix for `factor_layout`, set
  its position as if it were a latent
  factor.

- No need to specify the position of its
  indicators in the matrix for
  `factor_point_to`. Just leave the cell
  `NA`.

We generate the basic diagram by
`semPlot::semPaths()` first:

```{r}
library(semPlot)
library(semptools)
p <- semPaths(fit_sem,
              whatLabels = "est",
              sizeMan = 5,
              nCharNodes = 0,
              nCharEdges = 0,
              edge.width = 0.8,
              node.width = 0.7,
              edge.label.cex = 0.6,
              style = "ram",
              mar = c(10, 10, 10, 10),
              DoNotPlot = TRUE)
```

These are the vectors for `indicator_order`
and `indicator_factor`. Note that
`x04` and `x05` appear on both vectors.

```{r}
indicator_order  <- c("x04", "x05", "x01", "x02", "x03",
                      "x11", "x12", "x13", "x14", "x08", "x09", "x10")
indicator_factor <- c("x04", "x05", "f1",  "f1",  "f1",
                      "f4",  "f4",  "f4",  "f4",  "f3",  "f3",  "f3")
```

This is the matrix for `factor_layout`,
created by `layout_matrix()`:

```{r}
factor_layout <- layout_matrix(f1 = c(1, 1),
                               f3 = c(1, 2),
                               f4 = c(1, 3),
                               x04 = c(2, 1),
                               x05 = c(2, 2))
factor_layout
```

This is the matrix for `factor_point_to`,
created by `auto_factor_point_to()`
(added in `semptools` 0.2.9.15)
[^If not available, the matrix can be created manually by `matrix()`.]

```{r}
factor_point_to <- auto_factor_point_to(factor_layout,
                                        f1 = "up",
                                        f3 = "up",
                                        f4 = "up")
factor_point_to
```

Last, we created more vectors to further
modify the diagram:

```{r}
indicator_spread <- c(f4 = 1.5)
```

```{r}
p2 <- set_sem_layout(p,
                     indicator_order = indicator_order,
                     indicator_factor = indicator_factor,
                     factor_layout = factor_layout,
                     factor_point_to = factor_point_to,
                     indicator_spread = indicator_spread)
plot(p2)
```

# Final Remarks

For further information on using
`set_sem_layout()`, please refer to
`vignette("quick_start_sem")`. For
illustrations on how to modify a path
diagram from `semPlot::semPaths()`
using functions in `semptools`, please
refer to `vignette("semptools")`.